Particle Smoothing Algorithms for Variable Rate Models

### 2012-06-08 submit_doublecol_VRsmoothing.tex
Submitted for publication

\documentclass[journal]{IEEEtran}

@@

\author{Pete~Bunch,~\IEEEmembership{Member,~IEEE,}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

% The paper headers
\markboth{IEEE Transaction in Signal Processing,~Vol.~1, No.~1, January~1901}%
{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}

@@

\begin{abstract}
Standard state-space methods assume that the latent state evolves uniformly over time, and can be modelled with a discrete-time process synchronous with the observations. This may be a poor representation of some systems in which the state evolution displays discontinuities in its behaviour. For such cases, a variable rate model may be more appropriate; the system dynamics are conditioned on a set of random changepoints which constitute a marked point process. In this paper, new particle smoothing algorithms are presented for use with conditionally linear-Gaussian and conditionally deterministic dynamics. These are demonstrated on problems of financial modelling and target tracking. Results indicate that the smoothing approximations provide more accurate and more diverse representations of the state posterior distributions.
\end{abstract}

@@

The VRPF allows the changepoint sequence -- and hence the current state -- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable rate models, i.e. the estimation of the changepoint sequence and latent state given all the observations. We achieve this with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

We introduce the general structure of variable rate models in section~\ref{sec:vr_models}. The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}. In section~\ref{sec:simulations}, particular examples of variable rate models are presented and algorithm performance is demonstrated in a series of simulations.

@@

\begin{IEEEeqnarray}{rCl}
p(\theta_n^-) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
\end{IEEEeqnarray}

@@

This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.

@@

We first consider the class of variable rate models whose state dynamics are linear-Gaussian conditional on the changepoint sequence. Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM,
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A_n(\theta_{n}^-)x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n}^-)x_n + v_n       ,
\end{IEEEeqnarray}

@@

If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred conditionally using optimal Kalman filtering and smoothing recursions. As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step later in the paper.

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

The choice of weights determines the type of resampling used. The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}^-|y_{1:n-1})$. This, however, leads to degeneracy of the particle weights over time. Conventional resampling is achieved by using $v_{n-1}^{(j)} = w_{n-1}^{(j)}$. Any other choice results in an auxiliary particle filter \cite{Pitt1999}. For further discussion of resampling, see \cite{Cappe2007,Doucet2009,Douc2005}.

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_n)$, and concatenated with $\theta_{n-1}^{-(i)}$ to create a proposal for $\theta_n^-{(i)}$. Finally, the particle is weighted according to the ratio of the target and proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{-(i)}|y_{1:n}) }{ q(\theta_{n}^{-(i)}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) p(\theta_{n-1}^{-(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{-(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \label{eq:vrpf_weights}     .
\end{IEEEeqnarray}

@@

In \cite{Godsill2007}, the following choice of proposal weights was found to work well, as it preserves low probability particles which might turn out to be good estimates at later time steps,
%
\begin{IEEEeqnarray}{rCl}
v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )     .
\end{IEEEeqnarray}

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n^-, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}^-, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{-(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{-(i)})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_{n})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-backward method described in \cite{Godsill2004}.

@@

It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})$. The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations in variable rate models (i.e. $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}^-|\theta_{n}^{+}, y_{1:n}))$. However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable. This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}^-|\theta_{n}^+, y_{1:N})} \nonumber \\
\qquad & \propto & p(\theta_{n}^-, \theta_{n}^+| y_{1:N}) \nonumber  \\
       & =       & \int p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}^-, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}^-, y_{1:n}) dx_n \nonumber \\
       &   & \times p(\theta_{n}^+|\theta_{n}^-) p(\theta_{n}^-|y_{1:n})     .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}^-|y_{1:n})$ with weighted particles $\{\theta_{n}^{-(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{-(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots N_S$}
  	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
    \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
    \FOR{$n=N-1 \dots 1$}
      \STATE Discard $\theta_{n}^{-(i)}$.
      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict}).
      \FOR{$j=1 \dots N_F$}
	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
      \ENDFOR
      \STATE Sample $\theta_{n}^{-(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{-(j)}}(\theta_{n}^-)$.
      \STATE Join $\theta^{(i)} \gets \theta_{n}^{-(i)} \cup \theta_{n}^{+(i)}$.
      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using (\ref{eq:backward_kf_update}).
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised Variable Rate Particle Smoother}
\label{alg:RBVRPS}
\end{figure}

The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(N_F \times N_S \times N)$. It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, which avoids the necessity of calculating the sampling weights for every filter particle.

@@

In the conditionally deterministic models of section~\ref{sec:cd_models}, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$. When the changepoint sequence is split into past, $\theta_n^-$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future. This is undesirable in practice, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.

@@

The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}. Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^-, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability. Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.

>>>>

\documentclass[10pt,twocolumn,twoside]{IEEEtran}

@@

\author{Pete~Bunch*,~\IEEEmembership{Member,~IEEE,}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
}%\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

% The paper headers
\markboth{IEEE Transaction in Signal Processing}{IEEE Transaction in Signal Processing}%,~Vol.~1, No.~1, January~1901}%
%{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}

@@

\begin{abstract}
Standard state-space methods assume that the latent state evolves uniformly over time, and can be modelled with a discrete-time process synchronous with the observations. This may be a poor representation of some systems in which the state evolution displays discontinuities in its behaviour. For such cases, a variable rate model may be more appropriate; the system dynamics are conditioned on a set of random changepoints which constitute a marked point process. In this paper, new particle smoothing algorithms are presented for use with conditionally linear-Gaussian and conditionally deterministic dynamics. These are demonstrated on problems in financial modelling and target tracking. Results indicate that the smoothing approximations provide more accurate and more diverse representations of the state posterior distributions.
\end{abstract}

@@

The VRPF allows the changepoint sequence --- and hence the current state --- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable rate models, i.e. the estimation of the changepoint sequence and latent state given all the observations. We achieve this with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

We introduce the general structure of variable rate models in section~\ref{sec:vr_models}. The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}. In section~\ref{sec:simulations}, algorithm performance is demonstrated through simulations.

@@

\begin{IEEEeqnarray}{rCl}
p(\theta_n^-) & = & S(\tau_{K_n},u_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
\end{IEEEeqnarray}

@@

This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo (MCMC) algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.

@@

We first consider the class of variable rate models whose state dynamics are linear-Gaussian conditional on the changepoint sequence. Such a model may be discretised onto the set of observation times in exactly the same manner as a standard HMM,
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A_n(\theta_{n}^-)x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n}^-)x_n + v_n       ,
\end{IEEEeqnarray}

@@

If the changepoint sequence is known, or has been estimated, then the state posterior distributions may be inferred conditionally using optimal Kalman filtering and smoothing recursions. As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step later in the paper.

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-)} \nonumber \\
  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}+1},u_{K_{n-1}+1}, t_{n-1}) \nonumber \\
  & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

The choice of weights determines the type of resampling used. The simplest choice, $v_{n-1}^{(j)} = 1/\aleph_F$ (where $\aleph_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}^-|y_{1:n-1})$. This, however, leads to degeneracy of the particle weights over time. Conventional resampling is achieved by using $v_{n-1}^{(j)} = w_{n-1}^{(j)}$. Any other choice results in an auxiliary particle filter \cite{Pitt1999}. For further discussion of resampling, see \cite{Cappe2007,Doucet2009}.

Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_n)$, and concatenated with $\theta_{n-1}^{-(i)}$ to create a proposal for $\theta_n^{-(i)}$. Finally, the particle is weighted according to the ratio of the target and proposal densities,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{ p(\theta_{n}^{-(i)}|y_{1:n}) }{ q(\theta_{n}^{-(i)}) } \nonumber \\
    & \propto & \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) p(\theta_{n-1}^{-(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{-(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \nonumber \\
    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \label{eq:vrpf_weights}     .
\end{IEEEeqnarray}

@@

In \cite{Godsill2007}, the following choice of proposal weights was found to work well, as it preserves low probability particles which might turn out to be good estimates at later time steps,
%
\begin{IEEEeqnarray}{rCl}
v_{n-1}^{(i)} & \propto & \max ( 1, \aleph_F w_{n-1}^{(i)} )     .
\end{IEEEeqnarray}

@@

For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n^-, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter. Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations. It follows from the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
%
\begin{IEEEeqnarray}{rCl}
 p(x_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
 p(x_n|\theta_{n}^-, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
 p(y_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{-(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots \aleph_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{-(i)})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots \aleph_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_{n})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Piecewise deterministic variable rate particle filter}
\label{alg:VRPF}
\end{figure}

@@

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-backward method described in \cite{Godsill2004}.

@@

It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})$. The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations in variable rate models (i.e. $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}^-|\theta_{n}^{+}, y_{1:n}))$. However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable. This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle smoother for fixed rate conditionally linear-Gaussian models,
%
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}^-|\theta_{n}^+, y_{1:N})} \nonumber \\
\qquad & \propto & p(\theta_{n}^-, \theta_{n}^+| y_{1:N}) \nonumber  \\
       & =       & \int p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}^-, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}^-, y_{1:n}) dx_n \nonumber \\
       &   & \times p(\theta_{n}^+|\theta_{n}^-) p(\theta_{n}^-|y_{1:n})     .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}^-|y_{1:n})$ with weighted particles $\{\theta_{n}^{-(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{-(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots \aleph_S$}
  	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
    \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
    \FOR{$n=N-1 \dots 1$}
      \STATE Discard $\theta_{n}^{-(i)}$.
      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict}).
      \FOR{$j=1 \dots \aleph_F$}
	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
      \ENDFOR
      \STATE Sample $\theta_{n}^{-(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{-(j)}}(\theta_{n}^-)$.
      \STATE Join $\theta^{(i)} \gets \theta_{n}^{-(i)} \cup \theta_{n}^{+(i)}$.
      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using (\ref{eq:backward_kf_update}).
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised Variable Rate Particle Smoother}
\label{alg:RBVRPS}
\end{figure}

The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(\aleph_F \times \aleph_S \times N)$. It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, which avoids the necessity of calculating the sampling weights for every filter particle.

@@

In the conditionally deterministic models of section~\ref{sec:cd_models}, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$. When the changepoint sequence is split into past, $\theta_n^-$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future. This is undesirable in practice, as each sampling operation will require the recalculation of state values and likelihoods for all future observation times.

@@

The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}. Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^-, u_{K_n}\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability. Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.
